---
title: "FinalProj"
author: "Zoe Schubert"
date: "2024-04-14"
output: word_document
---

What vegetation microhabitat variables are the best determining factor for the presence of particular small mammal species?

```{r setup, include=FALSE}

library(ggplot2)
# install.packages("pander")
# install.packages("permutations")
# install.packages("gamlss.dist")
# install.packages("pscl")
# install.packages("gamlss")
# install.packages("vcd")
# install.packages("gtools")
# install.packages("devtools")
# install.packages("MuMIn")
# install.packages("permute")
# install.packages("MASS")
# install.packages("partitions")
library(interactions)
library(car)
library(lmtest)
library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)
library(gvlma)
library(rstatix)
library(tidyverse)
library(outliers)
library(psych)
library(caret)
library(report)
library(lubridate)
library(pscl)
library(glmtoolbox)

```

```{r}
setwd("C:/GitHub Projects/enec-562/Final Project")
surveydata <- read.csv("master.avg.csv")
```

```{r warning=FALSE}
library(dplyr)
#selecting columnns of interest - response variable and vegetation variables
filter_columns <- function(surveydata, columns_to_keep) {
  surveydata1 <- surveydata %>% select(columns_to_keep)
  return(surveydata1)
}

columns_to_keep <- c("total", "MaxHerb", "MaxShrub", "AvgDensity", "WoodyDebri", "GroundCover")

surveydata1 <- filter_columns(surveydata, columns_to_keep)

```

```{r echo = FALSE}
# histogram of total species counts
hist(surveydata1$total)
```
The frequency of species counts is right-skewed and zero-inflated, a poisson model will be used 

```{r}
# checking to see if the data are zero-inflated
100*sum(surveydata1$total == 0)/nrow(surveydata1)
```
```{r warning=FALSE}
library(gamlss.dist)
library(performance)
library(pscl)
survey_pois1 <-  glm(total ~ MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover, data = surveydata1, family="poisson")
check_zeroinflation(survey_pois1)
```
The data consist of 29.38% zeros, and checking for zero inflation from a Poisson model shows underfitting of zeros. Therefore, a zero-inflated model will be used for this analysis. 


Check dispersion:

```{r}
# checking overdispersion to decide if a neg-binomial should be used
E2 <- resid(survey_pois1, type = "pearson")
N  <- nrow(surveydata1)
p  <- length(coef(survey_pois1))  
sum(E2^2) / (N - p)
```
```{r}
#rootgram to validate overdispersion
library(gamlss)
library(vcd)
fit <- goodfit(survey_pois1$y, type = "pois") 
rootogram(fit)
```
A dispersion factor of 1.67, and the fit of the above rootgram indicate that overdispersion is not an issue here. Therefore, a zero-inflated Poisson Model can be used. 

Zero-inflated Poisson:
```{r}
library(gamlss.dist)
library(pscl)
zipois_model <- zeroinfl(total ~ MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover | ## Predictor for the Poisson process
                 MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover, ## Predictor for the Bernoulli process;
               dist = 'poisson',
               data = surveydata1)

summary(zipois_model)
```
Standard Poisson: 
```{r echo = FALSE, warning=FALSE}
# fitting the poisson
survey_pois <- glm(total ~ MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover, data = surveydata1, family="poisson")
summary(survey_pois)
```
Comparing mdoels with AIC and LR:

```{r warning=FALSE}

AIC(survey_pois, zipois_model)
lrtest(survey_pois, zipois_model)
```


Comparing models using prediction:

```{r}
#
r2_func <-function(preds,actual){ 
  return(1- sum((preds - actual) ^ 2)/sum((actual - mean(actual))^2))
}

RMSE_func <- function(preds, actual){
  return(sqrt(mean((actual - preds)^2)))
}
```


```{r warning=FALSE}
library(MASS)
set.seed(321)
compare_var <- as.data.frame(matrix(ncol = 3, nrow = 0))
colnames(compare_var) <- c("model_type", "R2", "RMSE")

Bootstrap_times <- 100 #this is how many times i want the loop to run
smp_size <- floor(0.70 * nrow(surveydata1)) #this is how i want to split training and testing data 


for(i in 1:Bootstrap_times) {
  train_ind <- sample(seq_len(nrow(surveydata1)), size = smp_size)
  train <- surveydata1[train_ind, ]
  test <- surveydata1[-train_ind, ]
  
  #models 
  pois_mod <- glm(total ~ MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover, data = train, family="poisson")
  nb_mod <- glm.nb(total ~ MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover, data = train)
  zipois_mod <- zeroinfl(total ~ MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover | ## Predictor for the Poisson process
                 MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover, ## Predictor for the Bernoulli process;
               dist = 'negbin',
               data = train)

  zinb_mod <- zeroinfl(total ~ MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover | ## Predictor for the Poisson process
                 MaxHerb + MaxShrub + AvgDensity + WoodyDebri + GroundCover, ## Predictor for the Bernoulli process;
               dist = 'poisson',
               data = train)

  
  #now predict the models on the test data 
  
  pois_pred <- predict(pois_mod, newdata = test, type = "response")
  nb_pred <- predict(nb_mod, newdata = test, type = "response")
  zipois_pred <- predict(zipois_mod, newdata = test, type = "response")
  zinb_pred <- predict(zinb_mod, newdata = test, type = "response")
 
  #make a dataframe for each modeling type
  pois <- as.data.frame(matrix(ncol =3))
  pois$V1 <- "pois"
  pois$V2 <- r2_func(pois_pred, test$total)
  pois$V3 <- RMSE_func(pois_pred, test$total)
  
  nb <- as.data.frame(matrix(ncol =3))
  nb$V1 <- "nb"
  nb$V2 <- r2_func(nb_pred, test$total)
  nb$V3 <- RMSE_func(nb_pred, test$total)
  
  zipois <- as.data.frame(matrix(ncol =3))
  zipois$V1 <- "zipois"
  zipois$V2 <- r2_func(zipois_pred, test$total)
  zipois$V3 <- RMSE_func(zipois_pred, test$total)
  
  zinb <- as.data.frame(matrix(ncol =3))
  zinb$V1 <- "zinb"
  zinb$V2 <- r2_func(zinb_pred, test$total)
  zinb$V3 <- RMSE_func(zinb_pred, test$total)
  
  #combine all of those 
  
  dat <- rbind(pois, nb, zipois, zinb)
colnames(dat) <- c("model_type", "R2", "RMSE")

  #combine it to our dataframe outside of the for loop 
  
  compare_var <- rbind(compare_var, dat)
}

#take the mean of all of those runs 

compare_var %>% pivot_longer(R2:RMSE, names_to = "metric", values_to = "values") %>% group_by(model_type, metric) %>% summarise(mean_val = mean(values)) %>% pivot_wider(names_from = "metric", values_from = "mean_val") %>% arrange(-R2)



```

According to the comparisons above, the zero-inflated Poisson model has a lower AIC, and the likelihood ratio test indicates a significant difference in the likelihood of the zero-inflated model and standard model in predicting total species counts. However, prediction indicates that the standard poisson has a lower RMSE than the zero-inflated poisson. We know that the model is zero-inflated, so the zero-inflated poisson should be used but may require cross-validation for predictor variables. 


Cross Validation: 
```{r}
# looping k-fold cross validation
library(jtools)
library(libcoin)
library(modelr)

library(MuMIn)
library(partitions)
library(permutations)
library(permute)
pastePerm<- function(row, names){
  keep<- which(row==1)
  if(length(keep)==0){
    return('1')
  }else{
    return(paste(names[keep],collapse='+'))
  }
}
my_sqrt <- function(var1){
  sqrt(var1)
}

dredgeform<- function(pred, covars, alwaysIn=''){
  p<- length(covars)
  perm.tab<- permutations(2, p, v=c(0,1), repeats.allowed=T)
  myforms<- NULL
  for(j in 1:nrow(perm.tab)){
    myforms[j]<- pastePerm(perm.tab[j,], covars)
  }
  myforms<- paste0(pred, '~ 1', alwaysIn,'+', myforms)
  return(myforms)
}

allformulas<- dredgeform(pred = "total", covars = c("MaxHerb", "MaxShrub", "AvgDensity", "WoodyDebri", "GroundCover"))
allformulas <- allformulas[2:length(allformulas)] 

```

```{r}
set.seed(123)
compare_var <- as.data.frame(matrix(ncol = 4, nrow = 0))
colnames(compare_var) <- c("formula", "RMSE", "R2", "MAE")

for ( i in 1:length(allformulas)) {
  
train.control <- trainControl(method = "repeatedcv", number = 3, repeats = 10)
# Train the full model
model_full <- train(as.formula(allformulas[i]), data = surveydata1, method = "pois",
               trControl = train.control)
# Summarize the results
compare_var[i, 1] <- allformulas[i]
compare_var[i, 2] <- mean(model_full$resample$RMSE)
compare_var[i, 3] <- mean(model_full$resample$Rsquared, na.rm = T)
compare_var[i, 4] <- mean(model_full$resample$MAE)


}

compare_var$prediction_error_rate <- compare_var$RMSE/mean(surveydata1$total) 

compare_var %>% arrange(prediction_error_rate)

compare_var %>% arrange(RMSE)
```








