---
title: "Mammal Vegetation Community Analysis"
author: "Nguyen Tien Anh Quach"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

This document is to analyze the vegetation-mammal community dataset. 

```{r message=F, warning=F, results=F}
library(ggplot2)
library(dplyr)
library(vegan)
library(rstatix)
library(ggpubr)

setwd("C:/GitHub Projects/enec-562/Final Project")
mam <- read.csv("master.avg.csv")

mam <- mam %>% filter(rowSums(select(.,7:13) != 0) > 0) %>%
    select(-avg.rara)

#some sites got 0, must excludec
```

# Select and check data

Select only count data: 

```{r message=F, warning=FALSE}
mam1 <- mam[,23:29] #exclude the columns that contains extra information that is not mammal count
```

Check for skewness and kurtosis:

```{r message=F, warning=F}
library(moments) #load moments package to check for kurtosis and skewness
skewness(mam1)
kurtosis(mam1)
```

As expect, count data are highly skewed. NMDS is the best!

## Plot the results

```{r message=F, warning=F}
anh_theme <- function() {
  theme(axis.text.x = element_text(hjust = 1)) +
    theme(axis.text.y = element_text(colour = "black", size = 10, face = "bold"), 
          axis.text.x = element_text(colour = "black", face = "bold", size = 10), 
          legend.text = element_text(size = 10, face ="bold", colour ="black"), 
          legend.position = "right", axis.title.y = element_text(face = "bold", size = 7), 
          axis.title.x = element_text(face = "bold", size = 12, colour = "black", margin = margin(t=5)), 
          axis.title.y.left = element_text(face = "bold", size = 12, colour = "black",margin=margin(r=5)),
          legend.title = element_text(size = 10, colour = "black", face = "bold"), 
          panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
          legend.key=element_blank(),
          plot.title = element_text(color = "black", size = 18, face = "bold", hjust = 0.5))
}
```

```{r message=F, warning=F}
#import data
setwd("C:/GitHub Projects/enec-562/Final Project")

library(readxl)
mam.scores <- read_excel("mam.scores.xlsx")
spp.scores <- as.data.frame(scores(mam_nmds,display=c("species")))

nmds_mam_plot <- ggplot(mam.scores, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 2,aes( color = Restored)) + # geom_polygon(data=data.scores.dplyr,aes(x=NMDS1,y=NMDS2,group=Subwatershed),alpha=0.30); this is to add a little convex polygon to visualize the clusters better. You can try to see how it looks. 
    scale_color_manual(values = c("red","green")) + anh_theme()
    # scale_shape_manual(values = c(18, 19,20,21,22,23,24, 25,26))

ggsave("NMDS plot.png", nmds_mam_plot, dpi = 300, width = 6, height = 4)

## Site NMDS

nmds.site_mam_plot <- ggplot(mam.scores, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 2,aes( color = Site)) + # geom_polygon(data=data.scores.dplyr,aes(x=NMDS1,y=NMDS2,group=Subwatershed),alpha=0.30); this is to add a little convex polygon to visualize the clusters better. You can try to see how it looks. 
  anh_theme()
    # scale_shape_manual(values = c(18, 19,20,21,22,23,24, 25,26))

ggsave("NMDS plot by site.png", nmds.site_mam_plot, dpi = 300, width = 6, height = 4)
```

```{r message=F, warning=F}
mam_biplot <- nmds_mam_plot + 
  geom_segment(data=spp.scores[c('avg.pele','avg.sihi',	'avg.rehu','avg.rara',
                                   'avg.orpa','avg.mumu','avg.mipi','avg.mipe'
),], aes(x=0, xend=NMDS1, y=0, yend=NMDS2), 
               color="black", arrow=arrow(length=unit(0.01,"npc"))) + #add arrow
  geom_text(data=spp.scores[c('avg.pele','avg.sihi',	'avg.rehu','avg.rara',
                                   'avg.orpa','avg.mumu','avg.mipi','avg.mipe'),], 
            aes(x=NMDS1,y=NMDS2,label=rownames(spp.scores[c('avg.pele','avg.sihi',	'avg.rehu','avg.rara',
                                   'avg.orpa','avg.mumu','avg.mipi','avg.mipe'),]),
                hjust=0.5*(1-sign(NMDS1)),vjust=0.5*(1-sign(NMDS2))), 
            color="black", size=4,fontface="bold")

ggsave("NMDS biplot.png", mam_biplot, dpi = 300, width = 6, height = 4)

```

From the look of the graph and investigation of data, restored sites seem to be very homogeneous, whereas there is much variability in composition at non-restored sites.

Let's try MRPP.

# MRPP of different groups

## Restored

```{r message=F, warning=F}

mam_restored <- mam1 %>% mutate(Restored = mam$Restored)
# Run MRPP
restored_mrpp <- mrpp(vegdist(mam_restored[,-8], method="bray"), mam_restored$Restored, 
                    distance = "bray", permutations = 999)

# Print the results
print(restored_mrpp)
```

MRPP, or Multi Response Permutation Procedure, is a statistical test used to assess whether there are significant differences between groups of sampling units based on a dissimilarity matrix calculated from multiple variables. MRPP is also great to use when data violate assumptions of normality or homogeneity of variances required by traditional parametric tests.
 
Results show that restored and non-restored sites have significantly different composition of mammals (p = 0.001).

## Site

```{r message=F, warning=F}

mam_site <- mam1 %>% mutate(Site = mam$Site)
# Run MRPP
site_mrpp <- mrpp(vegdist(mam_site[,-8], method="bray"), mam_site$Site, 
                    distance = "bray", permutations = 999)

# Print the results
print(site_mrpp)
```
Results show that traps at different sites have significantly different composition of mammals (p = 0.001).

# Community diversity indices comparison

## Shannon index

"The index takes into account the number of species living in a habitat (richness) and their relative abundance (evenness)."

```{r message=F, warning=F}
mam_shannon = as.data.frame(diversity(mam1, index = "shannon")) %>% 
  mutate(Restored = mam$Restored, Site = mam$Site)
head(mam_shannon)
```

 ### Test for Shannon index between restored and non-restored sites

```{r message=F, warning=F}
mam_shannon = mam_shannon %>% mutate(log.shannon = log(mam_shannon$`diversity(mam1, index = "shannon")`)) %>% rename("shannon" = `diversity(mam1, index = "shannon")`)

shapiro.test(mam_shannon$shannon)
shannon.test <- mam_shannon %>% wilcox_test(shannon ~ Restored) %>%
  add_significance()

mam_shannon %>% wilcox_effsize(shannon ~ Restored)

## plot

shannon.test <- shannon.test %>% add_xy_position(x = "Restored")
shannon_plot <- ggboxplot(
  mam_shannon, x = "Restored", y = "shannon", 
  ylab = "Shannon index", xlab = "Restoration status", add = "jitter"
  ) + 
  stat_pvalue_manual(shannon.test, tip.length = 0) +
  labs(subtitle = get_test_label(shannon.test, detailed = TRUE))

ggsave("Shannon plot.png", shannon_plot, dpi = 300, width = 6, height = 4)

```

## Simpson index

"The Shannon index stresses the richness component and rare cover types, whilst the Simpson index lays greater emphasis on the evenness component and on the dominant cover types (McGarigal and Marks, 1994, Haines-Young and Chopping, 1996, Riitters, Wickham, Vogelmann and Jones, 2000)."

Simpson gives more weight to dominant and common species.

```{r message=F, warning=F}
mam_simpson = as.data.frame(diversity(mam1, index = "simpson")) %>% 
  mutate(Restored = mam$Restored)
head(mam_simpson)
```

## Test for Simpson index difference

```{r message=F, warning=F}
mam_simpson = mam_simpson %>% rename("simpson" = `diversity(mam1, index = "simpson")`)
shapiro.test(mam_simpson$simpson)

simpson.test <- mam_simpson %>% wilcox_test(simpson ~ Restored) %>%
  add_significance()
mam_simpson %>% wilcox_effsize(simpson ~ Restored)

## plot

simpson.test <- simpson.test %>% add_xy_position(x = "Restored")
simpson_plot <- ggboxplot(
  mam_simpson, x = "Restored", y = "simpson", 
  ylab = "Simpson index", xlab = "Restoration status", add = "jitter"
  ) + 
  stat_pvalue_manual(shannon.test, tip.length = 0) +
  labs(subtitle = get_test_label(shannon.test, detailed = TRUE))

ggsave("Simpson plot.png", simpson_plot, dpi = 300, width = 6, height = 4)

```
