---
title: "Week 10 Problem Set"
author: "Nguyen Tien Anh Quach"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
chunk_output_type: inline
indent: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r, warning = FALSE, message = FALSE, results = F}
library(tidyverse)
library(tidycensus)
library(sf) 
library(tmap)
library(rmapshaper)
library(ggplot2)
library(spdep)
```

```{r, warning = FALSE, message = FALSE, results = F}
census_api_key("224b85597180fed0e9625fb771e54c72fa44490b", overwrite = TRUE, install = TRUE)
```

```{r, warning = FALSE, message = FALSE, results = F}
v18 <- load_variables(2018, "acs5", cache = TRUE)
```

```{r, warning = FALSE, message = FALSE, results = F}
nc.prob <- get_acs(geography = "tract", 
              year = 2018,
              variables = c(tpopr = "B03002_001",
                            nhwhite = "B03002_003", tot.bach = "B07009_005"), 
              state = "NC",
              survey = "acs5", 
              geometry = TRUE)

nc.prob = nc.prob %>% select(-moe) %>% rename (Tract = NAME) %>%
  pivot_wider(names_from = "variable", values_from = "estimate")

nc.prob <- nc.prob %>% mutate(pnhwhite = nhwhite/tpopr, ptot.bach = tot.bach/tpopr)
```

```{r, warning = FALSE, message = FALSE, results = F}
setwd("C:/GitHub Projects/enec-562/Week 10 - GIS")
NC_pollut <- read.csv("STATE_SINGLE_NC.csv")


#lets see how many of each type of site we have
NC_pollut %>% 
  group_by(SITE_TYPE_NAME) %>%
  summarise(no_rows = length(SITE_TYPE_NAME))

NC_pollut <- NC_pollut %>% filter(SITE_TYPE_NAME == "BROWNFIELDS SITE" | SITE_TYPE_NAME == "CONTAMINATED SITE" | SITE_TYPE_NAME == "POTENTIALLY CONTAMINATED SITE")

NC_pollut <- NC_pollut %>% filter(!is.na(LATITUDE83) & !is.na(LONGITUDE83))
NC_pollut_sf <- st_as_sf(NC_pollut, coords = c("LONGITUDE83", "LATITUDE83"), crs = 4269)

#make sure it has a coordinate system 
st_crs(NC_pollut_sf)
st_crs(nc.prob)

NC_pollut_sf_small <- st_join(x = NC_pollut_sf, y = nc.prob, 
                               join = st_within, left=FALSE)
```

```{r message=F, warning=F, fig.height = 5, fig.width = 10}
sites <- NC_pollut_sf_small %>% dplyr::select(geometry, INTEREST_TYPES, SITE_TYPE_NAME)
census_tracts_prob <- nc.prob

census_tracts_prob = st_transform(census_tracts_prob, crs = 5070)
sites = st_transform(sites, crs = 5070)

#plot

ggplot() +
  geom_sf(data = census_tracts_prob, aes(fill = pnhwhite), color = NA) + 
  labs(title = "Percent non-Hispanic white in North Carolina Tracts against hazardous waste sites") + 
  scale_fill_gradient(low = "#f7fcfd", high = "#00441b", na.value ="gray", name = "Percent non-Hispanic white") +  
  geom_sf(data = sites, aes(shape=SITE_TYPE_NAME)) +
  theme( axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())

```

```{r message=F, warning=F, fig.height = 5, fig.width = 10}
ggplot() +
  geom_sf(data = census_tracts_prob, aes(fill = ptot.bach), color = NA) + 
  labs(title = "Percent Bachelor's degree in North Carolina Tracts against hazardous waste sites") + 
  scale_fill_gradient(low = "#f7fcfd", high = "#bd0026", na.value ="gray", name = "Percent Bachelor's degree") +  
  geom_sf(data = sites, aes(shape=SITE_TYPE_NAME)) +
  theme( axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
```

2. Make two tables that summarize the census variables for areas within the buffer distance and outside of the buffer distance (mean, sd). 10

# Percent non Hispanic white within and outside 5 km buffer

```{r message=F, warning=F, results=F}
sel_nhwhite <- st_is_within_distance(census_tracts_prob, sites, dist = 5000)

summary(lengths(sel_nhwhite) > 0)

census_tracts_prox<- st_join(census_tracts_prob, sites,
            join = st_is_within_distance, dist = 5000)

census_tracts_prox$near_site <- ifelse(is.na(census_tracts_prox$INTEREST_TYPES), 0, 1)
census_tracts_prox <- census_tracts_prox %>% group_by(Tract) %>% summarize_all(mean, na.rm = TRUE)

#now lets rename that near_site variable 
census_tracts_prox$near_site <- ifelse(census_tracts_prox$near_site == 0, "far", "near")
```
The table below showed the mean and SD of percent non-Hispanic white within (near) and outside (far) of the 5 km buffer zone. 

```{r warning=F, message=F, echo=F}
library(kableExtra)
census_tracts_prox %>%
  st_drop_geometry() %>%
  filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  group_by(near_site) %>%
  summarise(mean.pnhwhite = mean(pnhwhite),
            sd.pnhwhite = sd(pnhwhite)) %>% 
  rename( "Proximity" = near_site,
          "Mean pct. non-Hispanic white" = mean.pnhwhite,
         "SD of pct. non-Hispanic white" = sd.pnhwhite) %>%
  kable() %>%
  column_spec(1:1, width = "1.5cm") %>%
  kable_styling(position = "center") %>%
  kable_styling(latex_options = "HOLD_position")
``` 

# Percent Bachelor's degree within and outside 5 km buffer

Similar to the table above, this table shows the percent Bachelor's degree of people living within (near) and outside (far) of the 5 km buffer. 

```{r warning=F, message=F, echo=F}
library(kableExtra)
census_tracts_prox %>%
  st_drop_geometry() %>%
  filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  group_by(near_site) %>%
  summarise(mean.ptotbach = mean(ptot.bach),
            sd.ptotbach = sd(ptot.bach)) %>% 
  rename( "Proximity" = near_site,
          "Mean pct. Bachelor" = mean.ptotbach,
         "SD of pct. Bachelor" = sd.ptotbach) %>%
  kable() %>%
  column_spec(1:1, width = "1.5cm") %>%
  kable_styling(position = "center") %>%
  kable_styling(latex_options = "HOLD_position")
```



3. Test the difference in means (or medians) for the two census variables inside and outside of the 5km buffer. Present your results from your tests in a few sentences. Make sure to check your assumptions. 20 points

# Testing near vs. far for percent non-Hispanic white

I used Shapiro-Wilk test to test for the normality of percent non-Hispanic white. The test yielded significant results (p < 0.001). I attempted to log transform the data, which did not help with the normality (p < 0.001). Therefore, I finally used the Wilcoxon test to test the difference in mean of percent non-Hispanic white between near and far tracts. 
\newline

The test yielded significant result (p < 0.001), showing that mean percent non-Hispanic white is significantly lower at near tracts (figure below).

```{r message=F, warning=F}
anh_theme <- function() {
  theme(axis.text.x = element_text(hjust = 1)) +
    theme(axis.text.y = element_text(colour = "black", size = 10, face = "bold"), 
          axis.text.x = element_text(colour = "black", face = "bold", size = 10), 
          legend.text = element_text(size = 10, face ="bold", colour ="black"), 
          legend.position = "right", axis.title.y = element_text(face = "bold", size = 7), 
          axis.title.x = element_text(face = "bold", size = 12, colour = "black", margin = margin(t=5)), 
          axis.title.y.left = element_text(face = "bold", size = 12, colour = "black",margin=margin(r=5)),
          legend.title = element_text(size = 10, colour = "black", face = "bold"), 
          panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
          legend.key=element_blank(),
          plot.title = element_text(color = "black", size = 18, face = "bold", hjust = 0.5))
}
```

```{r message=F, warning=F, results=F, fig.pos= 'center', fig.width=8, fig.height=6}
library(rstatix)
as.data.frame(census_tracts_prox) %>%
    filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  group_by(near_site) %>% shapiro_test(pnhwhite)
#not normally distributed

as.data.frame(census_tracts_prox) %>% 
    filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  mutate(log.pnhwhite = log(pnhwhite)) %>% group_by(near_site) %>% shapiro_test(log.pnhwhite)
#still not normal 

census_tracts_prox %>%   filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  wilcox_test(pnhwhite ~ near_site)

census_tracts_prox %>%   filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  ggplot() + geom_boxplot(aes(y = pnhwhite, x = near_site)) +
  anh_theme() + labs(x = "Proximity", y = "Percent non-Hispanic white")
```




# Testing near vs. far for percent Bachelor's degree

Similar to percent non-Hispanic white, percent Bachelor's degree data are non-normal before (p < 0.001) and after log-transformation (p < 0.001). Therefore, I used Wilcoxon test to test for the significant difference in mean percent Bachelor's degree between near and far sites.
\newline

The test yielded significant results (p < 0.001); therefore, percent people with Bachelor's degree at near sites is significantly lower than that at far sites (figure below).

```{r message=F, warning=F, results=F, fig.pos= 'center', fig.width=8, fig.height=6}
library(rstatix)
as.data.frame(census_tracts_prox) %>%   filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  group_by(near_site) %>% shapiro_test(ptot.bach)
#not normally distributed

as.data.frame(census_tracts_prox) %>%   filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  mutate(log.ptot.bach = log(ptot.bach)) %>% group_by(near_site) %>% shapiro_test(log.ptot.bach)
#still not normal 

census_tracts_prox %>%   filter(!is.na(pnhwhite) & !is.na(ptot.bach)) %>% 
  wilcox_test(ptot.bach ~ near_site)

census_tracts_prox %>% ggplot() + geom_boxplot(aes(y = ptot.bach, x = near_site)) +
  anh_theme() + labs(x = "Proximity", y = "Percent Bachelor's degree")
```


