---
title: "Data Visualization Basics"
author: "Sarah Roberts"
date: "Spring 2021"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## Opening discussion
Effective data visualization depends on purposeful choices about graph types. The ideal graph type depends on the type of data and the message the visualizer desires to communicate. The best visualizations are clear and simple. A good resource for data visualization is [Data to Viz](https://www.data-to-viz.com/), which includes both a decision tree for visualization types and explanation pages for each type of data, including links to R resources to create them. Take a few minutes to explore this website.

## Set Up

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)
#install.packages("ggridges")
library(ggridges)

PeterPaul.chem.nutrients <- 
  read.csv("./Data/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
PeterPaul.chem.nutrients.gathered <-
  read.csv("./Data/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv", stringsAsFactors = TRUE)
EPAair <- read.csv("./Data/EPAair_O3_PM25_NC1819_Processed.csv", stringsAsFactors = TRUE)

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
PeterPaul.chem.nutrients.gathered$sampledate <- as.Date(PeterPaul.chem.nutrients.gathered$sampledate, format = "%Y-%m-%d")

```

## ggplot

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_density_ridges
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot and Line Plot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot)

O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "blue")
print(O3plot2)

# Add additional variables
# How could you automatically assign a marker color to a variable? 
PMplot <- 
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year), color = Site.Name)) +
  geom_point()
print(PMplot)

# Separate plot with facets
PMplot.faceted <-
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3)
print(PMplot.faceted)

# Filter dataset within plot building and facet by multiple variables
PMplot.faceted2 <- 
  ggplot(subset(EPAair, Site.Name == "Clemmons Middle" | Site.Name == "Leggett" |
                  Site.Name == "Bryson City"), 
         aes(x = Month, y = PM2.5)) + 
  geom_point() +
  facet_grid(Site.Name ~ Year) 
print(PMplot.faceted2)

# Plot true time series with geom_line
PMplot.line <- 
  ggplot(subset(EPAair, Site.Name == "Leggett"), 
         aes(x = Date, y = PM2.5)) +
  geom_line()
print(PMplot.line)


```


### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +
  ylim(0, 20)
print(lightvsDOfixed)

# Depth in the fields of limnology and oceanography is on a reverse scale
tempvsdepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse()
print(tempvsdepth)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() +
  geom_smooth(method = lm) +
  geom_abline(aes(slope = 16, intercept = 0))
print(NvsP)

```

### Plotting continuous vs. categorical variables
A traditional way to display summary statistics of continuous variables is a bar plot with error bars. Let's explore why this might not be the most effective way to display this type of data. Navigate to the Caveats page on Data to Viz (https://www.data-to-viz.com/caveats.html) and find the page that explores barplots and error bars. 


```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"?
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient, fill = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge", dotsize = 2) #
print(Nutrientplot4)

# Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "darkred") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "blue") +
  geom_freqpoly(aes(x = no23), color = "royalblue") +
  geom_freqpoly(aes(x = po4), color = "red") 
print(Nutrientplot6)

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.chem.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)

# Frequency polygons have the risk of becoming spaghetti plots. 
# See <https://www.data-to-viz.com/caveat/spaghetti.html> for more info.

# Ridgeline plot
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(y = nutrient, x = concentration)) +
  geom_density_ridges(aes(fill = lakename), alpha = 0.5) #
print(Nutrientplot6)


```

### Themes
Often, we will want to change multiple visual aspects of a plot. Ggplot comes with pre-built themes that will adjust components of plots if you call that theme. 
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) 
print(O3plot)

O3plot1 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  theme_gray()
print(O3plot1)

O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  theme_bw()
print(O3plot2)

O3plot3 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  theme_light()
print(O3plot3)

O3plot4 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  theme_classic()
print(O3plot4)
```

Notice that some aspects of your graph have not been adjusted, including: 

* text size
* axis label colors
* legend position and justification

If you would like to set a common theme across all plots in your analysis session, you may define a theme and call up that theme for each graph. This eliminates the need to add multiple lines of code in each plot. 

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") #alternative: legend.position + legend.justification

# options: call the theme in each plot or set the theme at the start. 

O3plot5 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  mytheme
print(O3plot5)  

theme_set(mytheme)

O3plot6 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot6)  

```

### Adjusting multiple components of your plots

While the theme allows us to set multiple aspects of plots, ggplot allows us to adjust other parts of plots outside of the theme. 

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
O3plot7 <- ggplot(EPAair, aes(x = Date, y = Ozone)) +
  geom_hline(yintercept = 50, lty = 2) +
  geom_hline(yintercept = 100, lty = 2) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_text(x = as.Date("2020-01-01"), y = 45, label = "good", hjust = 1, fontface = "bold") +   # hjust = 1 means left align
  geom_text(x = as.Date("2020-01-01"), y = 95, label = "moderate", hjust = 1, fontface = "bold") +
  geom_text(x = as.Date("2020-01-01"), y = 120, label = "unhealthy (sensitive groups)", hjust = 1, fontface = "bold") +
  scale_x_date(limits = as.Date(c("2018-01-01", "2019-12-31")), 
    date_breaks = "2 months", date_labels = "%b %y") +
  ylab(expression("O"[3]* " AQI Value")) +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))
print(O3plot7)  

```

### Color palettes
Color palettes are an effective way to communicate additional aspects of our data, often illustrating a third categorical or continuous variable in addition to the variables on the x and y axes. A few rules for choosing colors: 

* Consider if your plot needs to be viewed in black and white. If so, choose a sequential palette with varying color intensity.
* Choose a palette that is color-blind friendly
* Maximize contrast (e.g., no pale colors on a white background)
* Diverging color palettes should be used for diverging values (e.g., warm-to-cool works well for values on a scale encompassing negative and positive values)

Does your color palette communicate additional and necessary information? If the answer is no, then you might consider removing it and going with a single color. Common instances of superfluous or redundant color palettes include: 

* Color that duplicates an axis
* Color that distinguishes categories when labels already exist (exception: if category colors repeat throughout a series of interrelated visualizations and help the reader build a frame of reference across a report)
* Color that reduces the conciseness of a plot

Perception is key! Choose palettes that are visually pleasing and will communicate what you are hoping your audience to perceive.

RColorBrewer (package)

* http://colorbrewer2.org
* https://moderndata.plot.ly/create-colorful-graphs-in-r-with-rcolorbrewer-and-plotly/

viridis and viridisLite (packages)

* https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
* https://ggplot2.tidyverse.org/reference/scale_viridis.html

colorRamp (function; comes with base R as part of the grDevices package)

* https://bookdown.org/rdpeng/exdata/plotting-and-color-in-r.html#colorramp

LaCroixColoR (package)

* https://github.com/johannesbjork/LaCroixColoR

wesanderson (package)

* https://github.com/karthik/wesanderson

nationalparkcolors (package)

* https://github.com/katiejolly/nationalparkcolors

```{r, warning = FALSE}
# install.packages("viridis")
# install.packages("RColorBrewer")
# install.packages("colormap")
library(viridis)
library(RColorBrewer)
library(colormap)

scales::show_col(colormap(colormap = colormaps$viridis, nshades = 16))
scales::show_col(colormap(colormap = colormaps$inferno, nshades = 16))
scales::show_col(colormap(colormap = colormaps$magma, nshades = 16))
display.brewer.all(n = 9)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth, shape = lakename)) +
  geom_point() 
print(NvsP)

# let's first make the plot look better.
# change your axis labels to reflect TN and TP in micrograms per liter.
# change your legend labels
NvsP2 <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth, shape = lakename)) +
  geom_point(alpha = 0.7, size = 2.5) +
  #labs() + # change your legend labels here
  scale_shape_manual(values = c(15, 17)) +
  #scale_color_distiller(palette = "Blues", direction = 1) + # use scale_color_brewer for discrete variables
  scale_color_viridis(option = "magma", direction = -1) +
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), legend.title = element_text(size = 12))
print(NvsP2)

# change your y axis label to list concentration in micrograms per liter
# remove your x axis label
# change labels for nutrients in the legend
# try out the different color palette options and choose one (or edit)
Nutrientplot <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration, color = nutrient)) +
  geom_boxplot() +
# place your additional edits here
  scale_y_continuous(expand = c(0, 0)) +
  #scale_color_brewer(palette = "YlGnBu") +
  #scale_color_manual(values = c("#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8", "#0c2c84")) +
  #scale_color_viridis(discrete = TRUE, end = 0.8) +
  theme(legend.position = "right")
print(Nutrientplot)
```


### Multiple plots on a page
In situations where facets don't fill our needs to place multiple plots on a page, we can use the package `cowplot` to arrange plots. The `plot_grid` function is extremely flexible in its ability to arrange plots in specific configurations. A useful guide can be found here: https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html.

A useful guide for aligning plots by axis can be found here: https://wilkelab.org/cowplot/articles/aligning_plots.html

```{r, fig.height = 7, fig.width = 6}
#install.packages("cowplot")
library(cowplot)
plot_grid(NvsP2, Nutrientplot, nrow = 2, align = 'h', rel_heights = c(1.25, 1))

```

### Saving plots
The `ggsave` function allows you to save plots in jpg, png, eps, pdf, tiff, and other formats. The following information can be supplied: 

* filename and relative path, with file extension and in quotes (required)
* plot object (required)
* width, height, units 
* resolution (dpi)

For example: 
`ggsave("./Output/PMplot.jpg", PMplot.faceted, height = 4, width = 6, units = "in", dpi = 300)`

## In class Assignment 

The following graph displays the counts of specific endpoints measured in neonicotinoid ecotoxicology studies. The way it is visualized, however, is not effective. Make the following coding changes to improve the graph and upload a screenshot of your code and figure: 

1. Change the ordering of the barplot so that the highest counts are listed first (left to right in descending order):   
2. Adjust the x axis labels so they appear at a 45 degree angle. 
3. Change the color and/or border on the bars. Make sure you have a different color for each bar.
```{r}
Neonics <- read.csv("./Data/ECOTOX_Neonicotinoids_Insects_raw.csv")
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint))

```

