---
title: "Data Wrangling"
author: "Sarah Roberts"
date: "Spring 2021"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Describe the usefulness of data wrangling and its place in the data pipeline
2. Wrangle datasets with dplyr functions
3. Apply data wrangling skills to a real-world example dataset

## Set up your session

Today we will work with a dataset from the [North Temperate Lakes Long-Term Ecological Research Station](https://lter.limnology.wisc.edu/about/overview). The NTL-LTER is located in the boreal zone in northern Wisconsin, USA. We will use the [chemical and physical limnology dataset](https://lter.limnology.wisc.edu/content/cascade-project-north-temperate-lakes-lter-core-data-physical-and-chemical-limnology-1984), running from 1984-2016. 


```{r, message = FALSE}
getwd()
#install.packages(tidyverse)
library(tidyverse)
#install.packages(lubridate)
library(lubridate)
detach("package:dplyr") #doing this makes it so stats doesn't interfere with your dplyr filter function
library(dplyr) 
library(plyr)
NTL.phys.data <- read.csv("Data/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

colnames(NTL.phys.data)
head(NTL.phys.data)
summary(NTL.phys.data)
str(NTL.phys.data)
dim(NTL.phys.data)

class(NTL.phys.data$sampledate)
# Format sampledate as date
NTL.phys.data$sampledate <- as.Date(NTL.phys.data$sampledate, format = "%m/%d/%y")
```

## Data Wrangling

Data wrangling extends data exploration: it allows you to process data in ways that are useful for you. An important part of data wrangling is creating *tidy datasets*, with the following rules: 

1. Each variable has its own column
2. Each observation has its own row
3. Each value has its own cell

What is the best way to wrangle data? There are multiple ways to arrive at a specific outcome in R, and we will illustrate some of those approaches. Your goal should be to write the simplest code that will get you to your desired outcome. However, there is sometimes a trade-off of the opportunity cost to learn a new formulation of code and the time it takes to write complex code that you already know. Remember that the best code is one that is easy to understand for yourself and your collaborators. Remember to comment your code, use informative names for variables and functions, and use reproducible methods to arrive at your output.

## Dplyr Wrangling Functions

`dplyr` is a package in R that includes functions for data manipulation (i.e., data wrangling or data munging). `dplyr` is included in the tidyverse package, so you should already have it installed on your machine. The functions act as verbs for data wrangling processes. For more information, run this line of code:

```{r, results = "hide"}
vignette("dplyr")
```

### Filter

Filtering allows us to choose certain rows (observations) in our dataset.

Here are the relevant commands used in the `filter` function. Add some notes to designate what these commands mean. 
`==`
`!=`
`<`
`<=`
`>`
`>=`
`&`
`|`

```{r}
class(NTL.phys.data$lakeid)
class(NTL.phys.data$depth)

# matrix filtering
NTL.phys.data.surface1 <- NTL.phys.data[NTL.phys.data$depth == 0,]

# dplyr filtering
NTL.phys.data.surface2 <- filter(NTL.phys.data, depth == 0)
NTL.phys.data.surface3 <- filter(NTL.phys.data, depth < 0.25)

# Did the methods arrive at the same result?
head(NTL.phys.data.surface1)
dim(NTL.phys.data.surface1)
head(NTL.phys.data.surface2)
dim(NTL.phys.data.surface2)
head(NTL.phys.data.surface3)
dim(NTL.phys.data.surface3)

# Choose multiple conditions to filter
summary(NTL.phys.data$lakename)
NTL.phys.data.PeterPaul1 <- filter(NTL.phys.data, lakename == "Paul Lake" | lakename == "Peter Lake")
NTL.phys.data.PeterPaul2 <- filter(NTL.phys.data, lakename != "Central Long Lake" & 
                                     lakename != "Crampton Lake" & lakename != "East Long Lake" &
                                     lakename != "Hummingbird Lake" & lakename != "Tuesday Lake" &
                                     lakename != "Ward Lake" & lakename != "West Long Lake")
NTL.phys.data.PeterPaul3 <- filter(NTL.phys.data, lakename %in% c("Paul Lake", "Peter Lake"))

# Choose a range of conditions of a numeric or integer variable
summary(NTL.phys.data$daynum)
NTL.phys.data.JunethruOctober1 <- filter(NTL.phys.data, daynum > 151 & daynum < 305)
NTL.phys.data.JunethruOctober2 <- filter(NTL.phys.data, daynum > 151, daynum < 305)
NTL.phys.data.JunethruOctober3 <- filter(NTL.phys.data, daynum >= 152 & daynum <= 304)
NTL.phys.data.JunethruOctober4 <- filter(NTL.phys.data, daynum %in% c(152:304))

# Exercise: 
# filter NTL.phys.data for the year 1999
# what code do you need to use, based on the class of the variable?
class(NTL.phys.data$year4)

# Exercise: 
# filter NTL.phys.data for Tuesday Lake from 1990 through 1999.

```
Question: Why don't we filter using row numbers?

> Answer: 

### Arrange

Arranging allows us to change the order of rows in our dataset. By default, the arrange function will arrange rows in ascending order.

```{r}
NTL.phys.data.depth.ascending <- arrange(NTL.phys.data, depth)
NTL.phys.data.depth.descending <- arrange(NTL.phys.data, desc(depth))

# Exercise: 
# Arrange NTL.phys.data by temperature, in descending order. 
# Which dates, lakes, and depths have the highest temperatures?


```
### Select

Selecting allows us to choose certain columns (variables) in our dataset.

```{r}
NTL.phys.data.temps <- select(NTL.phys.data, lakename, sampledate:temperature_C)

```
### Mutate

Mutating allows us to add new columns that are functions of existing columns. Operations include addition, subtraction, multiplication, division, log, and other functions.

```{r}

NTL.phys.data.temps <- mutate(NTL.phys.data.temps, temperature_F = (temperature_C*9/5) + 32)

```

## Lubridate

A package that makes coercing date much easier is `lubridate`. A guide to the package can be found at https://lubridate.tidyverse.org/. The cheat sheet within that web page is excellent too. This package can do many things (hint: look into this package if you are having unique date-type issues), but today we will be using two of its functions for our NTL dataset. 

```{r}
# add a month column to the dataset
NTL.phys.data.PeterPaul1 <- mutate(NTL.phys.data.PeterPaul1, month = month(sampledate)) 

# reorder columns to put month with the rest of the date variables
NTL.phys.data.PeterPaul1 <- select(NTL.phys.data.PeterPaul1, lakeid:daynum, month, sampledate:comments)

# find out the start and end dates of the dataset
interval(NTL.phys.data.PeterPaul1$sampledate[1], NTL.phys.data.PeterPaul1$sampledate[21613])
interval(first(NTL.phys.data.PeterPaul1$sampledate), last(NTL.phys.data.PeterPaul1$sampledate))
```


## Pipes

You all have used piping throughout the course, but it is good to keep practicing it. A good way to think about the function of a pipe is with the word "then." 

Let's say we want to take our raw dataset (NTL.phys.data), *then* filter the data for Peter and Paul lakes, *then* select temperature and observation information, and *then* add a column for temperature in Fahrenheit: 

```{r}
NTL.phys.data.processed <- 
  NTL.phys.data %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") %>%
  select(lakename, sampledate:temperature_C) %>%
  mutate(temperature_F = (temperature_C*9/5) + 32)
  
```

Notice that we did not place the dataset name inside the wrangling function but rather at the beginning.

### Saving processed datasets

```{r}
write.csv(NTL.phys.data.PeterPaul1, row.names = FALSE, file = "Data/NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv")
```

When we wrangle a raw dataset into a processed dataset, we create a code file that contains only the wrangling code. We then save the processed dataset as a new spreadsheet and then create a separate code file to analyze and visualize the dataset. Why do we keep the wrangling code separate from the analysis code?


Lets clear out our environment (note I am doing this so you don't get confused with all of the different data types, but it may not always be a good idea)
```{r}
rm(list = ls())
```

and load in our processed data from above
```{r}
NTL.phys.data.PeterPaul <- read.csv("./Data/NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv")
NTL.nutrient.data <- read.csv("./Data/NTL-LTER_Lake_Nutrients_Raw.csv",stringsAsFactors = TRUE)
```

## Review of basic exploration and wrangling
```{r}
# Data summaries for physical data
colnames(NTL.phys.data.PeterPaul)
dim(NTL.phys.data.PeterPaul)
str(NTL.phys.data.PeterPaul)
summary(NTL.phys.data.PeterPaul$comments)
class(NTL.phys.data.PeterPaul$sampledate)

# Format sampledate as date
NTL.phys.data.PeterPaul$sampledate <- as.Date(NTL.phys.data.PeterPaul$sampledate, format = "%Y-%m-%d")

# Select Peter and Paul Lakes from the nutrient dataset
NTL.nutrient.data.PeterPaul <- filter(NTL.nutrient.data, lakename == "Paul Lake" | lakename == "Peter Lake")

# Data summaries for nutrient data
colnames(NTL.nutrient.data.PeterPaul)
dim(NTL.nutrient.data.PeterPaul)
str(NTL.nutrient.data.PeterPaul)
summary(NTL.nutrient.data.PeterPaul$lakename) 
# Notice that other lake names didn't go away, even though they have zero values
NTL.nutrient.data.PeterPaul <- droplevels(NTL.nutrient.data.PeterPaul)
summary(NTL.nutrient.data.PeterPaul$lakename)
summary(NTL.nutrient.data.PeterPaul$comments)
class(NTL.nutrient.data.PeterPaul$sampledate)
NTL.nutrient.data.PeterPaul$sampledate <- as.Date(NTL.nutrient.data.PeterPaul$sampledate, format = "%m/%d/%y")

NTL.nutrient.data.PeterPaul <- 
  NTL.nutrient.data.PeterPaul %>% #
  mutate(month = month(sampledate)) %>% #
  select(lakeid:daynum, month, sampledate:comments) %>% # 
  drop_na(depth)

# Save processed nutrient file
write.csv(NTL.nutrient.data.PeterPaul, row.names = FALSE, 
          file = "./Data/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")

# Remove columns that are not of interest for analysis
NTL.phys.data.PeterPaul.subset <- select(NTL.phys.data.PeterPaul, 
                                         lakename:irradianceDeck)
  
NTL.nutrient.data.PeterPaul.subset <- select(NTL.nutrient.data.PeterPaul, 
                                             lakename, year4, daynum, month, sampledate, depth:po4)

```



## Gather and Spread

For most situations, data analysis works best when you have organized your data into a tidy dataset. A tidy dataset is defined as: 
* Each variable is a column
* Each row is an observation 
* Each value is in its own cell

However, there may be situations where we want to reshape our dataset, for example if we want to facet numerical data points by measurement type (more on this in the data visualization unit). We can program this reshaping in a few short lines of code using the package `tidyr`, which is conveniently included in the `tidyverse` package. 

We will work with the new functions from `tidyr`. For gather we will use `pivot_longer` and for spread we will use `pivot_wider`. Note that if you are familiar with `gather` and `spread` will may still use it, they are not going away, but they are not under active development.


```{r}
# Gather nutrient data into one column
#NTL.nutrient.data.PeterPaul.gathered <- gather(NTL.nutrient.data.PeterPaul.subset, "nutrient", "concentration", tn_ug:po4)

# Gather nutrient data into one column using pivot_longer
NTL.nutrient.data.PeterPaul.gathered <- pivot_longer(NTL.nutrient.data.PeterPaul.subset,tn_ug:po4, names_to = "nutrient", values_to = "concentration")

NTL.nutrient.data.PeterPaul.gathered <- subset(NTL.nutrient.data.PeterPaul.gathered, !is.na(concentration))
count(NTL.nutrient.data.PeterPaul.gathered, nutrient)

write.csv(NTL.nutrient.data.PeterPaul.gathered, row.names = FALSE, 
          file = "./Data/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")

# Spread nutrient data into separate columns
#NTL.nutrient.data.PeterPaul.spread <- spread(NTL.nutrient.data.PeterPaul.gathered, nutrient, concentration)

# Spread nutrient data into separate columns using pivot_wider
NTL.nutrient.data.PeterPaul.spread2 <- pivot_wider(NTL.nutrient.data.PeterPaul.gathered, names_from = nutrient, values_from = concentration)

# Split components of cells into multiple columns
# Opposite of 'separate' is 'unite'
NTL.nutrient.data.PeterPaul.dates <- separate(NTL.nutrient.data.PeterPaul.subset, sampledate, c("Y", "m", "d"))

# I recommend using lubridate rather than separate and unite.

```

## Combining multiple datasets

### Join 
In many cases, we will want to combine datasets into one dataset. If all column names match, the data frames can be combined with the `rbind` function. If some column names match and some column names don't match, we can combine the data frames using a "join" function according to common conditions that exist in the matching columns. We will demonstrate this with the NTL-LTER physical and nutrient datasets, where we have specific instances when physical and nutrient data were collected on the same date, at the same lake, and at the same depth. 

In dplyr, there are several types of join functions: 

* `inner_join`: return rows in x where there are matching values in y, and all columns in x and y (mutating join).
* `semi_join`: return all rows from x where there are matching values in  y, keeping just columns from x (filtering join).
* `left_join`: return all rows from x, and all columns from x and y (mutating join).
* `anti_join`: return all rows from x where there are *not* matching values in y, keeping just columns from x (filtering join).
* `full_join`: return all rows and all columns from x and y. Returns NA for missing values (mutating join).

Let's say we want to generate a new dataset that contains all possible physical and chemical data for Peter and Paul Lakes. In this case, we want to do a full join.

```{r}

NTL.phys.nutrient.data.PeterPaul <- full_join(NTL.phys.data.PeterPaul.subset, NTL.nutrient.data.PeterPaul.subset)

write.csv(NTL.phys.nutrient.data.PeterPaul, row.names = FALSE, 
          file ="./Data/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")

```

### rbind

The Niwot Ridge litter dataset, when downloaded from NEON, comes packaged with each month as a different .csv file. If we want to analyze the dataset as a single data frame, we need to combine each of these files. 

```{r}
Litter.June2016 <- read.csv("./Data/NIWO_Litter/NEON_NIWO_Litter_massdata_2016-06_raw.csv")
Litter.July2016 <- read.csv("./Data/NIWO_Litter/NEON_NIWO_Litter_massdata_2016-07_raw.csv")
Litter.August2016 <- read.csv("./Data/NIWO_Litter/NEON_NIWO_Litter_massdata_2016-08_raw.csv")

Litter.2019 <- rbind(Litter.June2016, Litter.July2016, Litter.August2016)
```

However, there are 20 months in this dataset, so importing all these files individually would be tedious to code. Here is a more efficient way to import and combine all files.

```{r}
LitterFiles = list.files(path = "./Data/NIWO_Litter/", pattern="*.csv", full.names=TRUE)
LitterFiles

Litter <- LitterFiles %>%
  ldply(read.csv)
```

We also have information about individual traps, including the location and type of landcover. Let's join these two datasets. Note that "siteID", "plotID" and "trapID" exist in both datasets, and we can join them by these conditions. Notice the dimensions of the final dataset.
```{r}
Trap <- read.csv("./Data/NEON_NIWO_Litter_trapdata_raw.csv")
LitterTrap <- left_join(Litter, Trap, by = c("siteID", "plotID", "trapID"))

dim(Litter)
dim(Trap)
dim(LitterTrap)

LitterTrap <- LitterTrap %>%
  select(plotID:trapID, collectDate, functionalGroup:qaDryMass, subplotID:geodeticDatum)

write.csv(LitterTrap, row.names = FALSE, 
          file ="./Data/NEON_NIWO_Litter_mass_trap_Processed.csv")
```

## Split-Apply-Combine

dplyr functionality, combined with the pipes operator, allows us to split datasets according to groupings (function: `group_by`), then run operations on those groupings and return the output of those operations. There is a lot of flexibility in this approach, but we will illustrate just one example today.

```{r}
NTL.PeterPaul.summaries <- 
  NTL.phys.nutrient.data.PeterPaul %>%
  filter(depth == 0) %>%
  group_by(lakename, month) %>%
  filter(!is.na(temperature_C) & !is.na(tn_ug) & !is.na(tp_ug)) %>%
  summarise(meantemp = mean(temperature_C), 
            sdtemp = sd(temperature_C), 
            meanTN = mean(tn_ug), 
            sdTN = sd(tn_ug), 
            meanTP = mean(tp_ug), 
            sdTP = sd(tp_ug))

write.csv(NTL.PeterPaul.summaries, row.names = FALSE, 
          file ="./Data/NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")

```

## Alternative Methods for Data Wrangling

If you want to iteratively perform operations on your data, there exist several options. We have demonstrated the pipe as one option. Additional options include the `apply` function (https://www.rdocumentation.org/packages/base/versions/3.5.2/topics/apply) and `for` loops (https://swcarpentry.github.io/r-novice-inflammation/15-supp-loops-in-depth/). These options are good options as well (again, multiple ways to get to the same outcome). A word of caution: loops are slow. This may not make a difference for small datasets, but small time additions will make a difference with large datasets.

